Re-analysing the data from @moffatt_inner_2020: What can we learn from an under-powered absence of difference?

### 2021-11-03 manuscript.Rmd
uploading revisions for collabra

---
title: "Re-analysing the data from @moffatt_inner_2020: A textbook illustration of the absence of evidence fallacy"

shorttitle: "The absence of evidence fallacy"

# note: \textcolor{red}{This is an example of a note}

author: 
  - name: "Ladislas Nalborczyk"
    affiliation: 1
    corresponding: yes
    address: "GIPSA-lab, CNRS, Univ. Grenoble Alpes, 11 Rue des Mathématiques, 38400 Saint-Martin-d'Hères, France"
    email: "ladislas.nalborczyk@univ-grenoble-alpes.fr"

affiliation:
  - id: 1
    institution: Univ. Grenoble Alpes, CNRS, Grenoble INP, GIPSA-lab, 38000 Grenoble, France

abstract: "@moffatt_inner_2020 reported the results of an experiment (N = 26 in the final sample) comparing the facial electromyographic correlates of mental rumination and distraction, following an experimentally induced stressor. Based on the absence of significant difference in the perioral muscular activity between the rumination and distraction conditions, @moffatt_inner_2020 concluded that *self-reported* inner experience was unrelated to peripheral muscular activity as assessed using surface electromyography. We suggest this conclusion is at best hasty. Indeed, concluding on the absence of an effect based on an under-powered non-significant *p*-value is strongly uninformative. Moreover, the relation between self-reports and physiological measures was not directly assessed, but only indirectly inferred from differences (or absence thereof) in group averages. Given the ample inter-individual variability in these measures (as suggested by our reanalysis), we think inferring the individual-level relation between self-reports and physiological measures from group averages is inappropriate. Given these limitations, we conclude that there is limited evidence for the main conclusion put forward by @moffatt_inner_2020 and we suggest ways forward, both from a theoretical and from a methodological perspective. Complete source code, reproducible analyses, and figures are available at https://osf.io/ba3gk/."

keywords: NHST, Bayesian, logical fallacy, reanalysis, inner speech, rumination, electromyography

csl: csl/apa7.csl
classoption: man, donotrepeattitle
linenumbers: yes
floatsintext: yes
link-citations: true
numbersections: false

bibliography: [bib/library.bib, bib/r-references.bib]

nocite: | 
  @R-base, @R-knitr, @R-rmarkdown, @R-tidyverse, @R-papaja,
  @R-wordcountaddin, @R-here

output:
  papaja::apa6_pdf:
    keep_tex: true
    highlight: tango
    latex_engine: xelatex
---

@@

# setting up knitr options
knitr::opts_chunk$set(
  cache = TRUE, eval = TRUE, echo = FALSE,
  warning = FALSE, message = FALSE,
  fig.align = "center", dev = "pdf",
  out.width = "100%", fig.width = 8,
  fig.pos = "!htb"
  )

@@

The activity of silently talking to oneself or "inner speech" is a foundational ability, allowing oneself to remember, plan, self-motivate or self-regulate [for reviews, see @alderson-day_inner_2015; @perrone-bertolotti_what_2014; @loevenbruck_cognitive_2018]. However, whereas the use of inner speech is associated with many adaptive functions in everyday life, inner speech dysfunctions can be identified in multiple psychological disorders. For instance, rumination, broadly defined as unconstructive repetitive thinking about past events and current mood states [@Martin], is involved in the onset and maintenance of serious mental disorders such as depression, anxiety, eating disorders, or substance abuse [for a review, see @Nolen-Hoeksema2008].

Given the predominantly verbal nature of rumination [e.g., @ehring_repetitive_2008; @goldwin_concreteness_2012; @goldwin_concreteness_2013; @mclaughlin_effects_2007], we previously proposed to consider rumination as a form of inner speech and to study it using the methods that have historically been used to study other forms of inner speech, namely, by using surface electromyography and motor interference protocols [e.g., @nalborczyk_orofacial_2017; @nalborczyk_understanding_2019; @nalborczyk_dissociating_2020; @nalborczyk_articulatory_2020]. We first showed that induced rumination was accompanied by increased facial (both over a forehead and a perioral site) muscular activity in comparison to a rest period [@nalborczyk_orofacial_2017]. However, because rumination was only compared to a rest period, it remained uncertain whether this perioral activity was specifically related to (inner) speech processes. Therefore, we ran a follow-up study comparing verbal to non-verbal rumination, which suggested that the facial EMG correlates we had previously identified were not specifically related to the verbal content of the ruminative thoughts [@nalborczyk_dissociating_2020]. We discussed these findings in length and proposed several theoretical interpretations that can account for these results in the discussion section of @nalborczyk_dissociating_2020 and more extensively in @nalborczyk_understanding_2019. Although these points were not addressed by @moffatt_inner_2020, their experimental design nevertheless had the potential to inform our understanding of the involvement of the speech motor system in different varieties of inner speech as well as to clarify the relation between the peripheral correlates of inner speech and (self-reported) subjective experience.

The main conclusion from @moffatt_inner_2020 is that inner experience between induced rumination and distraction differs "without a change in electromyographic correlates of inner speech" (p.1). In other words, they suggest that the subjective experience of inner speech is unrelated (or loosely related) to the electromyographic correlates of inner speech, which are thought to be represented mostly by the EMG amplitude recorded over the orbicularis oris inferior and orbicularis oris superior muscles. However, for this in-sample observation to be of interest in an out-of-sample context (i.e., to be informative for other non-observed individuals, or said otherwise, to bring information about the population), this absence of difference should be substantiated by adequately-powered statistical tests (given the target effect size) as well as reliable measures. This is unlikely to be the case here, for reasons that we will present and discuss in the present article. Moreover, a simple visual exploration of the data reveals important variability between individuals in the main effect of interest. That is, some participants had higher perioral muscular activity in the rumination condition than in the distraction condition, and some other participants showed the reverse pattern. This suggests unexplored variation in the determinants of this effect (e.g., the content of the inner experience). Indeed, the relation between the inner experience and the physiological correlates of inner speech production was only inferred from group averages. However, given the important inter-individual variability, this reasoning appears highly problematic. In the following, we explore each of these limitations and suggests ways forward, both from a theoretical and from a methodological perspective.

# Exploring the data

As typical in studies manipulating induced rumination, @moffatt_inner_2020 designed a two-step protocol. First, they aimed to induce a negative mood by asking participants to solve unsolvable or excessively difficult anagram and subtraction tasks. Second, they prompted participants to either ruminate on these (purportedly induced) negative feelings (by asking them to "think about the causes, consequences, and meaning of their current feelings") or to distract themselves (by asking them to "think about a village, city or town that you are particularly familiar with"). Rumination and distraction was manipulated within-subject, with all subjects alternating between rumination and distraction, in a counter-balanced order.

@@

Their final sample of participants, after data exclusion, included 26 participants (data available at https://osf.io/hj7tz/). The EMG data is depicted in Figure \@ref(fig:general) by condition (where BAS, DIS, and RUM refer to the baseline, distraction, and rumination conditions, respectively) and by muscle (frontalis, FRO; orbicularis oris inferior, OOI; and orbicularis oris superior, OOS). This figure shows that the average natural logarithm of the EMG peak amplitude recorded over the FRO was at similar levels in the baseline and distraction conditions, but was much higher in the rumination condition. However, the average natural logarithm of the EMG peak amplitude recorded over the OOI and OOS muscles was higher than baseline in both the rumination and distraction conditions, with a slight increase from distraction to rumination (both on the mean and median).

@@

# plotting the data
df2 %>%
    # reshapes data from wide to long format
    pivot_longer(
        cols = FRO:OOI,
        names_to = "muscle",
        values_to = "value",
        values_drop_na = TRUE
        ) %>%
    mutate(order = factor(order) ) %>%
    # plots the data
    ggplot(aes(x = condition, y = value) ) +
    # violin plots
    geom_violin(
        scale = "count",
        fill = "black",
        alpha = 0.15,
        colour = "white",
        position = pd,
        draw_quantiles = 0.5,
        show.legend = FALSE
        ) +
    # plotting individual data points
    geom_quasirandom(alpha = 0.3) +
    # plotting means
    stat_summary(
        fun = mean,
        geom = "line", size = 1,
        aes(group = muscle),
        position = pd,
        show.legend = FALSE
        ) +
    # plotting means
    stat_summary(
        fun = "mean",
        geom = "point", shape = 16, size = 3,
        position = pd,
        show.legend = TRUE
        ) +
    # plotting confidence intervals
    stat_summary(
        fun.data = mean_cl_normal,
        geom = "errorbar", size = 1, width = 0,
        fun.args = list(mult = 1.96),
        show.legend = FALSE,
        position = pd
        ) +
    # facetting by muscle
    facet_wrap(~muscle, ncol = 3) +
    # axis labels
    labs(x = "Condition", y = "Natural logarithm of peak amplitude") +
    theme_bw(base_size = 10) +
    scale_colour_brewer(palette = "Dark2", direction = 1) +
    scale_fill_brewer(palette = "Dark2", direction = 1)
```

@@

# fitting the model
constant_effects <- brm(
    mvbind(FRO, OOS, OOI) ~ 0 + Intercept + condition,
    family = gaussian(),
    prior = priors_constant,
    data = df2,
    chains = 4, cores = detectCores(),
    warmup = 2000, iter = 10000,
    control = list(adapt_delta = 0.95),
    sample_prior = TRUE,
    file = here("models/constant_effects")
    )

@@

# fitting the model
varying_effects <- brm(
    mvbind(FRO, OOS, OOI) ~ 0 + Intercept + condition + (0 + Intercept | ID),
    family = gaussian(),
    prior = priors_varying,
    data = df2,
    chains = 4, cores = detectCores(),
    warmup = 2000, iter = 10000,
    control = list(adapt_delta = 0.95),
    sample_prior = TRUE,
    file = here("models/varying_effects")
    )
```

```{r bf, eval = FALSE, fig.width = 10, fig.cap = "Savage-Dickey Bayes factor for the difference between the rumination and distraction conditions for each muscle. The BF is computed as the ratio of the posterior density to the prior density at $\\theta = 0$."}
# BF testing
hyp_fro <- hypothesis(
    varying_effects,
    "FRO_conditionDIS = FRO_conditionRUM"
    )

p1 <- plot(hyp_fro, plot = FALSE, theme = theme_bw(base_size = 12) )[[1]] +
    geom_vline(xintercept = 0, linetype = 2) +
    coord_cartesian(xlim = c(-5, 5) ) +
    theme_bw(base_size = 10) +
    theme(legend.title = element_blank() ) +
    labs(
      x = "Log EMG peak amplitude",
      title = paste0(
        "BF10 = ", round(1 / hyp_fro$hypothesis$Evid.Ratio, 3),
        " (BF01 = ", round(hyp_fro$hypothesis$Evid.Ratio, 3), ")"
        )
      )

@@

p2 <- plot(hyp_ooi, plot = FALSE, theme = theme_bw(base_size = 12) )[[1]] +
    geom_vline(xintercept = 0, linetype = 2) +
    coord_cartesian(xlim = c(-5, 5) ) +
    theme_bw(base_size = 10) +
    theme(legend.title = element_blank() ) +
    labs(
      x = "Log EMG peak amplitude",
      title = paste0(
        "BF10 = ", round(1 / hyp_ooi$hypothesis$Evid.Ratio, 3),
        " (BF01 = ", round(hyp_ooi$hypothesis$Evid.Ratio, 3), ")"
        )
      )

@@

p3 <- plot(hyp_oos, plot = FALSE, theme = theme_bw(base_size = 12) )[[1]] +
    geom_vline(xintercept = 0, linetype = 2) +
    coord_cartesian(xlim = c(-5, 5) ) +
    theme_bw(base_size = 10) +
    theme(legend.title = element_blank() ) +
    labs(
      x = "Log EMG peak amplitude",
      title = paste0(
        "BF10 = ", round(1 / hyp_oos$hypothesis$Evid.Ratio, 3),
        " (BF01 = ", round(hyp_oos$hypothesis$Evid.Ratio, 3), ")"
        )
      )

@@

Having nuanced some of the conclusions from @moffatt_inner_2020, we now turn to a discussion of the problems related to conclusions that can be made from under-powered non-significant results.

## Concluding on the null hypothesis from under-powered null-hypothesis significance tests: what could possibly go wrong?

There is an infamous tradition of conducting and interpreting uninformative null-hypothesis significance tests in Psychology [e.g., @harlow_problem_1997; @meehl_appraising_1990-1; @meehl_theoretical_1978; @meehl_theory-testing_1967; @meehl_why_1990]. By "uninformative", we mean that some null-hypothesis significance tests are simply not diagnostic with regards to the substantive effect of interest (e.g., whether there is a difference between conditions A and B).

@@

Anticipating the legitimate critiques on the power of their study, @moffatt_inner_2020 report the results of a power analysis using the effect size reported in @nalborczyk_orofacial_2017 of $d = 0.72$. This represents a highly optimistic estimate of the substantive effect of interest (i.e., the difference in the natural logarithm of the EMG peak amplitude between the rumination and distraction conditions) as this effect represents the standardised mean difference in EMG amplitude *between a rest and a rumination periods* [@nalborczyk_orofacial_2017].

@@

# plotting the results
parameter_grid %>%
  mutate(power = as.numeric(power) ) %>%
  ggplot(aes(x = sample_size, y = effect_size, colour = NULL) ) +
  geom_contour_filled(aes(z = power), alpha = 0.8) +
  # geom_raster(aes(fill = power), interpolate = TRUE) +
  geom_line(
    data = parameter_grid %>%
      filter(round(power, 2) == 0.9) %>%
      group_by(sample_size) %>%
      summarise(effect_size = mean(effect_size), power = mean(power) ) %>%
      ungroup,
    color = "black", lwd = 0.8
    ) +
  geom_segment(
    x = 26, xend = 26, y = 0, yend = target,
    lwd = 0.5, linetype = 3
    ) +
  geom_segment(
    x = 0, xend = 26, y = target, yend = target,
    lwd = 0.5, linetype = 3
    ) +
  geom_point(
    x = 26, y = target,
    pch = 21, size = 3, fill = "white"
    ) +
  theme_bw(base_size = 12) +
  labs(x = "Number of participants", y = "Standardised mean difference") +
  coord_cartesian(expand = FALSE) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank() ) +
  # scale_fill_manual(values = ., labels = seq(0.304, 2.129, length.out = 10) %>% round(2) )+
  # guides(fill = guide_colorbar() )
  # https://stackoverflow.com/questions/62543112/how-to-make-discrete-gradient-color-bar-with-geom-contour-filled
  # metR::scale_fill_discretised()
  # scale_fill_distiller(super = metR::ScaleDiscretised, palette = "RdYlBu") +
  guides(fill = guide_legend(title = "Power", reverse = TRUE) )
```

We suggest the (a priori) power of the study ran by @moffatt_inner_2020 was much lower than suggested by the authors. Indeed, we speculate that the standardised mean difference in EMG peak amplitude between the rumination and distraction conditions may be much weaker than the standardised mean difference in EMG amplitude between the rumination and rest conditions. If we assume that the former is half the size of the latter [which seems reasonable given the high inter-individual variability in such effects, cf. the next section but also @nalborczyk_can_2020], therefore the a priori power of the main statistical test from @moffatt_inner_2020 was around $0.42$, meaning that they had less than 1 chance out of 2 to find a significant effect (given that the population effect size was actually $0.36$). Notice that whereas taking half the effect size of @nalborczyk_orofacial_2017 may seem arbitrary, Figure \@ref(fig:power) shows that a one-sample t-test with a sample size of $N = 26$ is under-powered for a vast range of effect sizes.

@@

@@

While we appreciate the effort, the current approach poses new problems. First, contrary to what the authors suggest, whereas computing a BF indeed allows assessing the *relative evidence* for the null, computing a BF (i.e., comparing two models) does not solve the problem of low power. More precisely, the sensitivity (i.e., the ability to attain a certain goal) of an experimental design to detect a given effect is an issue for both frequentist and Bayesian statistical tests. To illustrate this point, we simulated 10.000 datasets (for $N = 26$) under the assumption of either no effect (i.e., the null hypothesis of $d = 0$), an effect size of $d = 0.36$ [i.e., the supposed target effect size in @moffatt_inner_2020], or an effect size of $d = 0.72$ [i.e., the effect size reported in @nalborczyk_orofacial_2017].

@@

```{r simulated-power, fig.cap = "Average natural logarithm of the Bayes factor in favour of the alternative hypothesis, along with its standard error, computed over 1000 datasets of increasing size simulated from the posterior predictive distribution of the varying-intercept multivariate Bayesian regresion model, fitted on the data from Moffatt et al. (2020). A log-BF belows 0 represents evidence for the null hypothesis (relative to the alternative hypothesis) and a log-BF above 0 represents evidence for the alternative hypothesis (relative to the null hypothesis)."}
# running the simulations
# source("code/bf_simulation.R")

>>>>

---
title: "Re-analysing the data from @moffatt_inner_2020: What can we learn from an under-powered absence of difference?"
shorttitle: "The absence of evidence fallacy"
# note: \textcolor{red}{This is an example of a note}
author: 
  - name: "Ladislas Nalborczyk"
    affiliation: "1, 2, 3"
    corresponding: yes
    address: "LPC, CNRS, Aix-Marseille University, 13 Boulevard Gustave Desplaces, 13003 Marseille, France"
    email: "ladislas.nalborczyk@univ-amu.fr"
affiliation:
  - id: 1
    institution: Univ. Grenoble Alpes, CNRS, Grenoble INP, GIPSA-lab, 38000 Grenoble, France
  - id: 2
    institution: Aix Marseille Univ, CNRS, LPC, Marseille, France
  - id: 3
    institution: Aix Marseille Univ, CNRS, LNC, Marseille, France
abstract: "@moffatt_inner_2020 reported the results of an experiment (N = 26 in the final sample) comparing the facial electromyographic correlates of mental rumination and distraction, following an experimentally induced stressor. Based on the absence of significant difference in the perioral muscular activity between the rumination and distraction conditions, @moffatt_inner_2020 concluded that *self-reported* inner experience was unrelated to peripheral muscular activity as assessed using surface electromyography. We suggest this conclusion is at best hasty. Indeed, concluding on the absence of an effect based on an under-powered non-significant *p*-value is strongly uninformative. In this short commentary, we show that there is limited evidence for the main conclusion put forward by @moffatt_inner_2020. We suggest ways forward, both from a theoretical and methodological perspective. Complete source code, reproducible analyses, and figures are available at https://osf.io/ba3gk/."
keywords: NHST, Bayes factor, reanalysis, inner speech, rumination, electromyography
csl: csl/apa7.csl
classoption: man, donotrepeattitle
linenumbers: yes
floatsintext: yes
link-citations: true
numbersections: false
mask: yes # masking author names
bibliography: [bib/library.bib, bib/r-references.bib]
nocite: | 
  @R-base, @R-knitr, @R-rmarkdown, @R-tidyverse, @R-papaja,
  @R-wordcountaddin, @R-here
output:
  papaja::apa6_pdf:
    keep_tex: true
    highlight: tango
    latex_engine: xelatex
editor_options: 
  chunk_output_type: console
---

@@

# setting up knitr options
knitr::opts_chunk$set(
  cache = TRUE, eval = TRUE, echo = FALSE,
  warning = FALSE, message = FALSE,
  fig.align = "center", dev = "cairo_pdf",
  out.width = "100%", fig.width = 8,
  fig.pos = "!htb"
  )

@@

The activity of silently talking to oneself or "inner speech" is a foundational ability, allowing oneself to remember, plan, self-motivate, or self-regulate [for reviews, see @alderson-day_inner_2015; @perrone-bertolotti_what_2014; @loevenbruck_cognitive_2018]. However, whereas the use of inner speech is associated with many adaptive functions in everyday life, inner speech dysfunctions can be identified in multiple psychological disorders. For instance, rumination, broadly defined as unconstructive repetitive thinking about past events and current mood states [@Martin], is involved in the onset and maintenance of serious mental disorders such as depression, anxiety, eating disorders, or substance abuse [for review, see @Nolen-Hoeksema2008].

Given the predominantly verbal nature of rumination [e.g., @ehring_repetitive_2008; @goldwin_concreteness_2012; @goldwin_concreteness_2013; @mclaughlin_effects_2007], we previously proposed to consider rumination as a form of inner speech and to study it using the methods that have historically been used to study other forms of inner speech, namely, by using surface electromyography (EMG) and motor interference protocols [e.g., @nalborczyk_dissociating_2021; @nalborczyk_orofacial_2017; @nalborczyk_understanding_2019; @nalborczyk_articulatory_2020]. We first showed that induced rumination was accompanied by increased facial (both over a forehead and a perioral site) muscular activity in comparison to a rest period [@nalborczyk_orofacial_2017]. However, because rumination was only compared to a rest period, it remained uncertain whether this perioral activity was specifically related to (inner) speech processes. Therefore, we ran a follow-up study comparing verbal to non-verbal rumination, which suggested that the facial EMG correlates we had previously identified were not specifically related to the verbal content of the ruminative thoughts [@nalborczyk_dissociating_2021]. We discussed these findings in length and proposed several theoretical interpretations that can account for these results in the discussion of @nalborczyk_dissociating_2021 and more extensively in @nalborczyk_understanding_2019.

Although these points were not addressed by @moffatt_inner_2020, their experimental design nevertheless had the potential to inform our understanding of the involvement of the speech motor system in different varieties of inner speech as well as to clarify the relation between the peripheral correlates of inner speech and (self-reported) subjective experience.

@moffatt_inner_2020 designed an experiment with the aim of refining our understanding of the involvement of the speech motor system in different varieties of inner speech and clarifying the relation between the peripheral correlates of inner speech and (self-reported) subjective experience. Their main conclusion is that inner experience between induced rumination and distraction differs "without a change in electromyographic correlates of inner speech" (p.1). In other words, they suggest that the subjective experience of inner speech is unrelated (or loosely related) to the electromyographic correlates of inner speech, which are thought to be represented mostly by the EMG amplitude recorded over the orbicularis oris inferior and orbicularis oris superior muscles. However, for this in-sample observation to be of interest in an out-of-sample context (i.e., to be informative for other non-observed individuals, or said otherwise, to bring information about the population), this absence of difference should be substantiated by adequately-powered statistical tests (given the target effect size) as well as reliable measures. This is unlikely to be the case here, for reasons that we will present and discuss in the present article.

Moreover, a simple visual exploration of the data reveals important variability between individuals in the main effect of interest. That is, some participants had higher perioral muscular activity in the rumination condition than in the distraction condition, and some other participants showed the reverse pattern. This suggests unexplored variation in the determinants of this effect (e.g., the content of the inner experience). In the following, we explore each of these limitations and suggests ways forward, both from a theoretical and from a methodological perspective.

Indeed, the relation between the inner experience and the physiological correlates of inner speech production was only inferred from group averages. However, given the important inter-individual variability, this reasoning appears highly problematic

# Re-analysing the data from @moffatt_inner_2020

## Exploring the data

As typical in studies manipulating induced rumination, @moffatt_inner_2020 designed a two-step protocol. First, they aimed to induce a negative mood by asking participants to solve unsolvable or excessively difficult anagram and subtraction tasks. Second, they prompted participants to either ruminate on these (purportedly induced) negative feelings (by asking them to "think about the causes, consequences, and meaning of their current feelings") or to distract themselves (by asking them to "think about a village, city or town that you are particularly familiar with"). Rumination and distraction were manipulated within-subject, with all subjects alternating between rumination and distraction, in a counter-balanced order.

@@

Their final sample of participants, after data exclusion, included 26 participants (data available at https://osf.io/hj7tz/). The EMG data is depicted in Figure \@ref(fig:general) by condition (where BAS, DIS, and RUM refer to the baseline, distraction, and rumination conditions, respectively) and by muscle (frontalis, FRO; orbicularis oris inferior, OOI; and orbicularis oris superior, OOS). This figure shows that the average natural logarithm of the EMG peak amplitude recorded over the FRO was at similar levels in the baseline and distraction conditions, but was much higher in the rumination condition. However, the average natural logarithm of the EMG peak amplitude recorded over the OOI and OOS muscles was higher than baseline in both the rumination and distraction conditions, with a slight increase from distraction to rumination (both on the mean and median). Having described the data collected by @moffatt_inner_2020, we now turn to a discussion of some problems related to conclusions that can be made from under-powered non-significant results.

@@

# plotting the data
df2 %>%
    # reshapes data from wide to long format
    pivot_longer(
        cols = FRO:OOI,
        names_to = "muscle",
        values_to = "value",
        values_drop_na = TRUE
        ) %>%
    mutate(order = factor(order) ) %>%
    # plots the data
    ggplot(aes(x = condition, y = value) ) +
    # violin plots
    geom_violin(
        scale = "count",
        fill = "black",
        alpha = 0.15,
        colour = "white",
        position = pd,
        draw_quantiles = 0.5,
        show.legend = FALSE
        ) +
    # plotting individual data points
    geom_quasirandom(alpha = 0.3) +
    # plotting means
    stat_summary(
        fun = mean,
        geom = "line", size = 1,
        aes(group = muscle),
        position = pd,
        show.legend = FALSE
        ) +
    # plotting means
    stat_summary(
        fun = "mean",
        geom = "point", shape = 16, size = 3,
        position = pd,
        show.legend = TRUE
        ) +
    # plotting 95% confidence intervals
    stat_summary(
        fun.data = mean_cl_normal,
        geom = "errorbar", size = 1, width = 0,
        fun.args = list(mult = 1.96),
        show.legend = FALSE,
        position = pd
        ) +
    # facetting by muscle
    facet_wrap(~muscle, ncol = 3) +
    # axis labels
    labs(x = "Condition", y = "Natural logarithm of peak amplitude") +
    theme_bw(base_size = 12, base_family = "Open Sans") +
    scale_colour_brewer(palette = "Dark2", direction = 1) +
    scale_fill_brewer(palette = "Dark2", direction = 1)
```

@@

# fitting the model
constant_effects <- brm(
    mvbind(FRO, OOS, OOI) ~ 0 + Intercept + condition,
    family = gaussian(),
    prior = priors_constant,
    data = df2,
    chains = 4, cores = 4,
    warmup = 2000, iter = 10000,
    control = list(adapt_delta = 0.95),
    sample_prior = TRUE,
    file = here("models/constant_effects")
    )

@@

# fitting the model
varying_effects <- brm(
    mvbind(FRO, OOS, OOI) ~ 0 + Intercept + condition + (0 + Intercept | ID),
    family = gaussian(),
    prior = priors_varying,
    data = df2,
    chains = 4, cores = 4,
    warmup = 2000, iter = 10000,
    control = list(adapt_delta = 0.95),
    sample_prior = TRUE,
    file = here("models/varying_effects")
    )
```

```{r bf, eval = FALSE, fig.width = 10, fig.cap = "Savage-Dickey Bayes factor for the difference between the rumination and distraction conditions for each muscle. The BF is computed as the ratio of the posterior density to the prior density at $\\theta = 0$."}
# BF testing
hyp_fro <- hypothesis(
    x = varying_effects,
    hypothesis = "FRO_conditionDIS = FRO_conditionRUM"
    )

p1 <- plot(hyp_fro, plot = FALSE, theme = theme_bw(base_size = 12) )[[1]] +
    geom_vline(xintercept = 0, linetype = 2) +
    coord_cartesian(xlim = c(-5, 5), ylim = c(0, 7.5) ) +
    theme_bw(base_size = 12, base_family = "Open Sans") +
    theme(legend.title = element_blank() ) +
    labs(
      x = "Log EMG peak amplitude",
      y = "Probability density",
      title = paste0(
        "BF10 = ", round(1 / hyp_fro$hypothesis$Evid.Ratio, 3),
        " (BF01 = ", round(hyp_fro$hypothesis$Evid.Ratio, 3), ")"
        )
      )

@@

p2 <- plot(hyp_ooi, plot = FALSE, theme = theme_bw(base_size = 12) )[[1]] +
    geom_vline(xintercept = 0, linetype = 2) +
    coord_cartesian(xlim = c(-5, 5), ylim = c(0, 7.5) ) +
    theme_bw(base_size = 12, base_family = "Open Sans") +
    theme(legend.title = element_blank() ) +
    labs(
      x = "Log EMG peak amplitude",
      title = paste0(
        "BF10 = ", round(1 / hyp_ooi$hypothesis$Evid.Ratio, 3),
        " (BF01 = ", round(hyp_ooi$hypothesis$Evid.Ratio, 3), ")"
        )
      )

@@

p3 <- plot(hyp_oos, plot = FALSE, theme = theme_bw(base_size = 12) )[[1]] +
    geom_vline(xintercept = 0, linetype = 2) +
    coord_cartesian(xlim = c(-5, 5), ylim = c(0, 7.5) ) +
    theme_bw(base_size = 12, base_family = "Open Sans") +
    theme(legend.title = element_blank() ) +
    labs(
      x = "Log EMG peak amplitude",
      title = paste0(
        "BF10 = ", round(1 / hyp_oos$hypothesis$Evid.Ratio, 3),
        " (BF01 = ", round(hyp_oos$hypothesis$Evid.Ratio, 3), ")"
        )
      )

@@

## Conclusions from under-powered null-hypothesis significance tests

There is an infamous tradition of conducting and interpreting uninformative null-hypothesis significance tests in Psychology [e.g., @meehl_appraising_1990-1; @harlow_problem_1997; @meehl_theoretical_1978; @meehl_theory-testing_1967; @meehl_why_1990]. By "uninformative", we mean that some null-hypothesis significance tests are simply not diagnostic with regards to the substantive effect of interest (e.g., whether there is a difference between conditions A and B).

@@

## Enthusiastic prior specifications

Anticipating the legitimate critiques on the power of their study, @moffatt_inner_2020 report the results of a power analysis using the effect size reported in @nalborczyk_orofacial_2017 of $d = 0.72$. This represents a highly optimistic estimate of the substantive effect of interest (i.e., the difference in the natural logarithm of the EMG peak amplitude between the rumination and distraction conditions) as this effect represents the standardised mean difference in EMG amplitude *between a rest and a rumination periods* as estimated in @nalborczyk_orofacial_2017.

@@

# plotting the results
parameter_grid %>%
  mutate(power = as.numeric(power) ) %>%
  ggplot(aes(x = sample_size, y = effect_size, colour = NULL) ) +
  geom_contour_filled(aes(z = power), alpha = 0.8) +
  # geom_raster(aes(fill = power), interpolate = TRUE) +
  geom_line(
    data = parameter_grid %>%
      filter(round(power, 2) == 0.9) %>%
      group_by(sample_size) %>%
      summarise(effect_size = mean(effect_size), power = mean(power) ) %>%
      ungroup,
    color = "black", lwd = 0.8
    ) +
  geom_segment(
    x = 26, xend = 26, y = 0, yend = target,
    lwd = 0.5, linetype = 3
    ) +
  geom_segment(
    x = 0, xend = 26, y = target, yend = target,
    lwd = 0.5, linetype = 3
    ) +
  geom_point(
    x = 26, y = target,
    pch = 21, size = 3, fill = "white"
    ) +
  theme_bw(base_size = 12, base_family = "Open Sans") +
  labs(x = "Number of participants", y = "Standardised mean difference") +
  coord_cartesian(expand = FALSE) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank() ) +
  # scale_fill_manual(values = ., labels = seq(0.304, 2.129, length.out = 10) %>% round(2) )+
  # guides(fill = guide_colorbar() )
  # https://stackoverflow.com/questions/62543112/how-to-make-discrete-gradient-color-bar-with-geom-contour-filled
  # metR::scale_fill_discretised()
  # scale_fill_distiller(super = metR::ScaleDiscretised, palette = "RdYlBu") +
  guides(fill = guide_legend(title = "Power", reverse = TRUE) )
```

```{r power2, eval = FALSE, fig.cap = "Statistical power as a function of both sample size and effect size, for a one-sample t-test with a significance level of 0.05. The white dot indicates the minimal effect size that can be detected with a probability equal or superior to 0.9 with a sample size of N = 26."}
# creating a grid of possible sample and effect sizes
parameter_grid <- expand.grid(
  alpha = seq.int(from = 0, to = 0.2, length.out = 100),
  effect_size = seq.int(from = 0.1, to = 1, length.out = 300)
  )

# computing power for all combinations of sample and effect sizes
parameter_grid$power <- power.t.test(
  n = 26,
  delta = parameter_grid$effect_size,
  sd = 1,
  type = "one.sample",
  alternative = "two.sided",
  sig.level = parameter_grid$alpha)$power

# computing the effect size for a power = 0.8 at sample size = 26
target <- parameter_grid %>%
  filter(round(alpha, 2) == 0.05, round(power, 2) == 0.9) %>%
  pull(effect_size) %>%
  mean

# plotting the results
parameter_grid %>%
  mutate(power = as.numeric(power) ) %>%
  ggplot(aes(x = alpha, y = effect_size, colour = NULL) ) +
  geom_contour_filled(aes(z = power), alpha = 0.8) +
  geom_line(
    data = parameter_grid %>%
      filter(round(power, 2) == 0.9) %>%
      group_by(alpha) %>%
      summarise(effect_size = mean(effect_size), power = mean(power) ) %>%
      ungroup,
    color = "black", lwd = 0.8
    ) +
  geom_segment(
    x = 0.05, xend = 0.05, y = 0, yend = target,
    lwd = 0.5, linetype = 3
    ) +
  geom_segment(
    x = 0, xend = 0.05, y = target, yend = target,
    lwd = 0.5, linetype = 3
    ) +
  geom_point(
    x = 0.05, y = target,
    pch = 21, size = 3, fill = "white"
    ) +
  theme_bw(base_size = 12, base_family = "Open Sans") +
  labs(x = "Alpha level", y = "Standardised mean difference") +
  coord_cartesian(expand = FALSE) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank() ) +
  # scale_fill_manual(values = ., labels = seq(0.304, 2.129, length.out = 10) %>% round(2) )+
  # guides(fill = guide_colorbar() )
  # https://stackoverflow.com/questions/62543112/how-to-make-discrete-gradient-color-bar-with-geom-contour-filled
  # metR::scale_fill_discretised()
  # scale_fill_distiller(super = metR::ScaleDiscretised, palette = "RdYlBu") +
  guides(fill = guide_legend(title = "Power", reverse = TRUE) )
```

We suggest the (a priori) power of the study ran by @moffatt_inner_2020 was much lower than suggested by the authors. Indeed, we speculate that the standardised mean difference in EMG peak amplitude between the rumination and distraction conditions may be much weaker than the standardised mean difference in EMG amplitude between the rumination and rest conditions. If we assume that the former is half the size of the latter, therefore the a priori power of the main statistical test from @moffatt_inner_2020 was around $0.42$, meaning that they had less than 1 chance out of 2 to find a significant effect (given that the population effect size was actually $0.36$). Notice that whereas taking half the effect size of @nalborczyk_orofacial_2017 may seem arbitrary, Figure \@ref(fig:power) shows that a one-sample t-test with a sample size of $N = 26$ is under-powered for a vast range of effect sizes.

@@

## Frequentist properties of Bayes factors

@@

However, the current approach poses new problems. First, contrary to what the authors suggest, whereas computing a BF indeed allows assessing the *relative evidence* for the null, computing a BF (i.e., comparing two models) does not solve the problem of low power. More precisely, the sensitivity (i.e., the ability to attain a certain goal) of an experimental design to detect a given effect is an issue for both frequentist and Bayesian statistical tests. To illustrate this point, we simulated 10.000 datasets (for $N = 26$) under the assumption of either no effect (i.e., the null hypothesis of $d = 0$), an effect size of $d = 0.36$ [i.e., the supposed target effect size in @moffatt_inner_2020], or an effect size of $d = 0.72$ [i.e., the effect size reported in @nalborczyk_orofacial_2017].

@@

```{r simulated-power, eval = FALSE, fig.cap = "Average natural logarithm of the Bayes factor in favour of the alternative hypothesis, along with its standard error, computed over 1000 datasets of increasing size simulated from the posterior predictive distribution of the varying-intercept multivariate Bayesian regresion model, fitted on the data from Moffatt et al. (2020). A log-BF belows 0 represents evidence for the null hypothesis (relative to the alternative hypothesis) and a log-BF above 0 represents evidence for the alternative hypothesis (relative to the null hypothesis)."}
# running the simulations
# source("code/bf_simulation.R")
